---
title: "Fakinumab IV bolus — linear and MM clearance"
classoption: openany
output:
  rmarkdown::html_vignette:
    toc: true
bibliography: references.bib
biblio-style: apalike
link-citations: yes
vignette: >
  %\VignetteIndexEntry{Fakinumab IV bolus — linear and MM clearance}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r global_options, echo = FALSE, include = FALSE}
knitr::opts_knit$set(tangle = FALSE)
options(width = 120)
knitr::opts_chunk$set(
  purl = FALSE,
  collapse = TRUE, comment = "#>", echo = FALSE,
  warning = FALSE, message = FALSE, cache = FALSE
)
library(PFIM)
library(ggplot2)
```

# Overview

This vignette reproduces the PopED **mrgsolve** tutorial on *fakinumab* first-in-human
PK: a two-compartment model with **linear** and **Michaelis–Menten (MM)** elimination
from the central compartment, six parallel IV bolus dose groups, and sparse PK sampling
over 21 days.

PopED uses `mrgsim_q()` inside the `ff()` wrapper; PFIM implements the same structural
model as a custom ODE (`CENT`, `PERIPH` amounts; observation `CP = CENT/V1`).

Interactive workflow:

```r
devtools::load_all(".")
source("tools/run_fakinumab_evaluation.R")
```

# Study

| Item | Value |
|------|--------|
| Route | Single IV bolus |
| Dose groups (mg) | 0.03, 0.1, 0.3, 1, 3, 10 |
| Subjects per group | 6 |
| Sampling (days) | 1 h, 4 h, 1, 3, 7, 14, 21 |

PopED passes dose as `DOSE` (mg) and multiplies by 1000 in the simulation data set
(central amount in µg). PFIM uses the same convention: `dose = DOSE_mg × 1000` on
`CENT`.

# Structural model

mrgsolve / PopED (`model_poped.cpp`):

```text
ke  = CL/V1;  k12 = Q/V1;  k21 = Q/V2;  CP = CENT/V1
dCENT/dt  = k21*PERIPH - k12*CENT - VMAX*CP/(KM+CP) - ke*CENT
dPERIPH/dt = k12*CENT - k21*PERIPH
```

PFIM ODE (equivalent):

```{r model-eq, echo = TRUE, eval = FALSE}
modelEquations = list(
  Deriv_CENT = paste(
    "{ CP = CENT/V1; ke = CL/V1; k12 = Q/V1; k21 = Q/V2;",
    "k21*PERIPH - k12*CENT - VMAX*CP/(KM + CP) - ke*CENT }"
  ),
  Deriv_PERIPH = "{ k12 = Q/V1; k21 = Q/V2; k12*CENT - k21*PERIPH }"
)
outputs = list(RespPK = "CENT/V1")
```

# Population parameters and residual error

PopED `bpop`, IIV (`d`), and proportional error (`feps.prop`, `sigma = 0.15`):

```{r params, echo = TRUE, eval = FALSE}
modelParameters = list(
  ModelParameter(name = "CL",   distribution = LogNormal(mu = 0.5,  omega = 0.2)),
  ModelParameter(name = "VMAX", distribution = LogNormal(mu = 20,   omega = 0.2)),
  ModelParameter(name = "KM",   distribution = LogNormal(mu = 1.2,  omega = 0)),
  ModelParameter(name = "V1",   distribution = LogNormal(mu = 2.5,  omega = 0.1)),
  ModelParameter(name = "Q",    distribution = LogNormal(mu = 10,   omega = 0)),
  ModelParameter(name = "V2",   distribution = LogNormal(mu = 4,   omega = 0))
)

modelError = list(
  Proportional(output = "RespPK", sigmaSlope = 0.15)
)
```

Log-normal IIV on `CL`, `VMAX`, and `V1` only (`omega` = 0 on `KM`, `Q`, `V2`), matching
PopED `d = c(CL = 0.2, VMAX = 0.2, V1 = 0.1)`.

# Design

```{r design, echo = TRUE, eval = FALSE}
sample_days = c(1 / 24, 4 / 24, 1, 3, 7, 14, 21)
doses_mg    = c(0.03, 0.1, 0.3, 1, 3, 10)

fakinumab_arm = function(dose_mg) {
  Arm(
    name = paste(dose_mg, "mg"),
    size = 6,
    administrations = list(
      Administration(
        outcome = "CENT",
        timeDose = 0,
        dose = dose_mg * 1000
      )
    ),
    samplingTimes = list(
      SamplingTimes(outcome = "RespPK", samplings = sample_days)
    ),
    initialConditions = list(CENT = 0, PERIPH = 0)
  )
}

design_fakinumab = Design(
  name = "fakinumab",
  arms = lapply(doses_mg, fakinumab_arm)
)

evaluation_fakinumab = Evaluation(
  name = "Fakinumab PopED mrgsolve",
  modelEquations = modelEquations,
  modelParameters = modelParameters,
  modelError = modelError,
  outputs = outputs,
  designs = list(design_fakinumab),
  fimType = "population",
  odeSolverParameters = list(atol = 1e-8, rtol = 1e-8)
)
```

```{r defs, include = FALSE}
eval(parse(text = knitr::knit_code$get("model-eq")))
eval(parse(text = knitr::knit_code$get("params")))
eval(parse(text = knitr::knit_code$get("design")))
```

# Test plot (population mean)

```{r test-plot, echo = TRUE, fig.width = 8, fig.height = 5}
plot_fakinumab_test = function(evaluation, n_points = 200) {
  model = defineModelWrapper(defineModelType(evaluation), evaluation)
  arms  = S7::prop(S7::prop(evaluation, "designs")[[1]], "arms")
  times = seq(0, 21, length.out = n_points)

  curves = purrr::list_rbind(purrr::map(seq_along(arms), function(i) {
    plot_arm = arms[[i]]
    S7::prop(plot_arm, "samplingTimes") = list(
      SamplingTimes("RespPK", times)
    )
    m = defineModelAdministration(model, plot_arm)
    y = evaluateModel(m, plot_arm)
    data.frame(
      time = y$RespPK$time,
      CP = y$RespPK$RespPK,
      dose = S7::prop(arms[[i]], "name")
    )
  }))

  ggplot(curves, aes(time, CP, color = dose)) +
    geom_line(linewidth = 0.9) +
    scale_y_log10(limits = c(0.01, 1e4)) +
    labs(
      x = "Time from dose (days)",
      y = expression(C[P] ~ "(mg/L, log scale)"),
      color = "Dose group",
      title = "Fakinumab — population mean PK by dose"
    )
}

plot_fakinumab_test(evaluation_fakinumab)
```

# Population FIM and RSE

```{r fim, echo = TRUE, eval = FALSE}
evaluation_fakinumab = run(evaluation_fakinumab)
stopifnot(getDeterminant(evaluation_fakinumab) > 0)
show(evaluation_fakinumab)
getRSE(evaluation_fakinumab)
```

```{r run-fim, include = FALSE}
evaluation_fakinumab = run(evaluation_fakinumab)
```

Reference RSE (%) from PopED `get_rse(evaluate.fim(poped_db_mrg), poped_db_mrg)`:

```{r compare-poped, echo = TRUE}
poped_rse = c(
  mu_CL = 10.87, mu_VMAX = 10.66, mu_KM = 7.94,
  mu_V1 = 6.36, mu_Q = 8.89, mu_V2 = 3.62,
  w2_CL = 38.99, w2_VMAX = 32.10, w2_V1 = 31.67,
  sigma = 10.78
)

rse = getRSE(evaluation_fakinumab)
pfim_rse = c(
  mu_CL = rse$RSE[match("μ_CL", rownames(rse))],
  mu_VMAX = rse$RSE[match("μ_VMAX", rownames(rse))],
  mu_KM = rse$RSE[match("μ_KM", rownames(rse))],
  mu_V1 = rse$RSE[match("μ_V1", rownames(rse))],
  mu_Q = rse$RSE[match("μ_Q", rownames(rse))],
  mu_V2 = rse$RSE[match("μ_V2", rownames(rse))],
  w2_CL = rse$RSE[match("ω²_CL", rownames(rse))],
  w2_VMAX = rse$RSE[match("ω²_VMAX", rownames(rse))],
  w2_V1 = rse$RSE[match("ω²_V1", rownames(rse))],
  sigma = rse$RSE[match("σ_slope_RespPK", rownames(rse))]
)

cmp = data.frame(
  parameter = names(poped_rse),
  PopED = unname(poped_rse),
  PFIM = unname(pfim_rse),
  rel_diff_pct = 100 * abs(pfim_rse - poped_rse) / poped_rse,
  stringsAsFactors = FALSE
)
cmp_num = cmp
cmp_num$parameter = NULL
cmp = cbind(cmp["parameter"], round(cmp_num, 2))
knitr::kable(cmp, caption = "RSE (%): PopED vs PFIM")
```

**Interpretation:** the design is well identified in both tools (`det(FIM) > 0`). Fixed
effects and IIV RSE are **same order of magnitude**; remaining differences come from
PopED’s mrgsolve-based likelihood linearisation vs PFIM’s finite-difference gradients
on the ODE, and from how very small doses (0.03 mg) interact with MM clearance at the
numerical tolerance.

# Sampling windows (PopED)

PopED evaluates window efficiency with `plot_efficiency_of_windows()` (±1 h on day 1,
±3 h on later samples). In PFIM, use `SamplingTimeConstraints` on each arm and re-run
`run()` or an optimization algorithm to explore local sampling windows.

# Related vignettes

- `vignette("Example01")`, `vignette("Example02")` — design evaluation and optimization
- `vignette("Example03_GCSF")` — larger PK/PD ODE aligned with PopED
- `vignette("LibraryOfModels")` — built-in 2-compartment MM models (MM-only elimination)
