---
title: "G-CSF PK/PD design evaluation"
classoption: openany
output:
  rmarkdown::html_vignette:
    toc: true
bibliography: references.bib
biblio-style: apalike
link-citations: yes
vignette: >
  %\VignetteIndexEntry{G-CSF PK/PD design evaluation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r global_options, echo = FALSE, include = FALSE}
knitr::opts_knit$set(tangle = FALSE)
options(width = 120)
knitr::opts_chunk$set(
  purl = FALSE,
  collapse = TRUE, comment = "#>", echo = FALSE,
  warning = FALSE, message = FALSE, cache = FALSE
)
library(PFIM)
library(ggplot2)
```

# Overview

Filgrastim (G-CSF) PK/PD design evaluation with PFIM: three parallel dose groups
(1, 3, 10 mg/kg SC every 24 h for 7 doses, body weight 75 kg, 10 subjects per group),
dense PK (`CP`) and ANC (`NB`) sampling over 10 days [@Krzyzanski2010].

To reproduce this workflow interactively in RStudio:

1. Set the working directory to the **PFIM package root** (folder with `DESCRIPTION`).
2. Run `devtools::load_all(".")` — not `library(PFIM)` alone (that loads an older installed build).
3. **Do not** re-use an `evaluation_gcsf` object from a previous session. Either run all
   code chunks below in order, or in one step:

```r
devtools::load_all(".")
source("tools/run_gcsf_evaluation.R")  # creates evaluation_gcsf + runs FIM
```

**Quick check:** after `run()`, `det(FIM) > 0` and `getRSE(evaluation_gcsf)["μ_NB0", "RSE"]`
should be about **11%**, not ~1000%. If you only see 8 fixed effects (`μ_KA`…`μ_SM2` without
`μ_KD`, `μ_KINT`, `μ_KMT`, `μ_KTT`), you are still on the old parameterization.

The hidden `defs` chunk runs automatically when knitting only.

# Model equations

```{r model-eq, echo = TRUE, eval = FALSE}
# Baseline steady-state block (MAIN): derived from primary parameters so that
# gradients w.r.t. NB0, KMT, etc. stay consistent (as in mrgsolve/PopED).
gcsf_main_block = paste(
  "BAS = 0.0246; CP0 = BAS;",
  "H10 = 1+SM1*CP0/(SC1+CP0); H20 = 1+SM2*CP0/(SC1+CP0); H30 = 1+SM3*CP0/(SC1+CP0);",
  "KINB = KMT*NB0/H10;",
  "den = KBB1*H30 + KTT*H20;",
  "BM10 = KINB*H10/den; BM20 = KTT*H20*BM10/den;",
  "BM30 = KTT*H20*BM20/den; BM40 = KTT*H20*BM30/den;",
  "BM50 = KTT*H20*BM40/den; BM60 = KTT*H20*BM50/den;",
  "BM70 = KTT*H20*BM60/den; BM80 = KTT*H20*BM70/den; BM90 = KTT*H20*BM80/den;",
  "NT0 = BM10+BM20+BM30+BM40+BM50+BM60+BM70+BM80+BM90+NB0;",
  "AC0 = CP0*VD; ADR0 = KSI*NT0*AC0/(KD+CP0);",
  "KIN = KEL*AC0 + KINT*ADR0;"
)

gcsf_ode_prefix = paste(
  gcsf_main_block,
  "ATOT = CENT;",
  "NT = B1+B2+B3+B4+B5+B6+B7+B8+B9+NB; RTOT = KSI*NT;",
  "BB = RTOT-ATOT/VD+KD; CP = 0.5*(-BB+sqrt(BB^2+4*KD*ATOT/VD));",
  "AC = CP*VD; ADR = RTOT*AC/(KD+CP);",
  "H1 = 1+SM1*CP/(SC1+CP); H2 = 1+SM2*CP/(SC1+CP); H3 = 1+SM3*CP/(SC1+CP);"
)

modelEquations = list(
  Deriv_ABS  = paste0("{ ", gcsf_ode_prefix, " -KA*ABS }"),
  Deriv_CENT = "{ KA*ABS + KIN - KEL*AC - KINT*ADR }",
  Deriv_FRR  = "0", Deriv_DR = "0",
  Deriv_B1   = "{ KINB*H1*(B1/BM10) - KBB1*H3*B1 - KTT*H2*B1 }",
  Deriv_B2   = "{ KTT*H2*B1 - KBB1*H3*B2 - KTT*H2*B2 }",
  Deriv_B3   = "{ KTT*H2*B2 - KBB1*H3*B3 - KTT*H2*B3 }",
  Deriv_B4   = "{ KTT*H2*B3 - KBB1*H3*B4 - KTT*H2*B4 }",
  Deriv_B5   = "{ KTT*H2*B4 - KBB1*H3*B5 - KTT*H2*B5 }",
  Deriv_B6   = "{ KTT*H2*B5 - KBB1*H3*B6 - KTT*H2*B6 }",
  Deriv_B7   = "{ KTT*H2*B6 - KBB1*H3*B7 - KTT*H2*B7 }",
  Deriv_B8   = "{ KTT*H2*B7 - KBB1*H3*B8 - KTT*H2*B8 }",
  Deriv_B9   = "{ KTT*H2*B8 - KBB1*H3*B9 - KTT*H2*B9 }",
  Deriv_NB   = "{ KTT*H2*B9 + KBB1*H3*(B1+B2+B3+B4+B5+B6+B7+B8+B9) - KMT*NB }"
)

outputs = list(
  RespPK = paste0(
    "0.5*(-(KSI*(B1+B2+B3+B4+B5+B6+B7+B8+B9+NB)-CENT/VD+KD)",
    "+sqrt((KSI*(B1+B2+B3+B4+B5+B6+B7+B8+B9+NB)-CENT/VD+KD)^2+4*KD*CENT/VD))"
  ),
  RespPD = "NB"
)
```

# Parameters, error model, initial conditions

```{r params, echo = TRUE, eval = FALSE}
# DDMORE population mean (Krzyzanski et al.)
p = c(FF=0.626, KA=0.642, KINT=0.101, KSI=0.211, KMT=0.0723, KBB1=0, KTT=0.0102,
       KEL=0.148, VD=2.56, KD=1.27, NB0=1.65, SC1=3.21, SM1=34.3, SM2=32.3, SM3=0)

# Primary parameters only (KINB, BM10, KIN are derived in gcsf_main_block).
# KD, KINT, KMT, KTT are estimated fixed effects in PopED (no IIV).
modelParameters = list(
  ModelParameter(name="KA",   distribution=LogNormal(mu=p["KA"],   omega=0)),
  ModelParameter(name="KEL",  distribution=LogNormal(mu=p["KEL"],  omega=sqrt(0.312))),
  ModelParameter(name="VD",   distribution=LogNormal(mu=p["VD"],   omega=sqrt(0.328))),
  ModelParameter(name="KD",   distribution=LogNormal(mu=p["KD"],   omega=0)),
  ModelParameter(name="KINT", distribution=LogNormal(mu=p["KINT"], omega=0)),
  ModelParameter(name="KSI",  distribution=LogNormal(mu=p["KSI"],  omega=sqrt(0.224))),
  ModelParameter(name="KMT",  distribution=LogNormal(mu=p["KMT"],  omega=0)),
  ModelParameter(name="KTT",  distribution=LogNormal(mu=p["KTT"],  omega=0)),
  ModelParameter(name="NB0",  distribution=LogNormal(mu=p["NB0"],  omega=sqrt(0.298))),
  ModelParameter(name="SC1",  distribution=LogNormal(mu=p["SC1"],  omega=sqrt(0.803))),
  ModelParameter(name="SM1",  distribution=LogNormal(mu=p["SM1"],  omega=sqrt(0.0128))),
  ModelParameter(name="SM2",  distribution=LogNormal(mu=p["SM2"],  omega=0)),
  ModelParameter(name="SM3",  distribution=LogNormal(mu=p["SM3"],  omega=0), fixedMu=TRUE),
  ModelParameter(name="FF",   distribution=LogNormal(mu=p["FF"],   omega=0), fixedMu=TRUE),
  ModelParameter(name="KBB1", distribution=LogNormal(mu=p["KBB1"],omega=0), fixedMu=TRUE)
)

modelError = list(
  Combined1(output="RespPK", sigmaInter=0,    sigmaSlope=0.253),
  Combined1(output="RespPD", sigmaInter=2.10, sigmaSlope=0.0227)
)

# String expressions: re-evaluated from current mu on each gradient step.
gcsf_baseline_ic = function(state) {
  blk = paste(
    "local({ BAS = 0.0246; CP0 = BAS;",
    "H10 = 1+SM1*CP0/(SC1+CP0); H20 = 1+SM2*CP0/(SC1+CP0); H30 = 1+SM3*CP0/(SC1+CP0);",
    "KINB = KMT*NB0/H10; den = KBB1*H30 + KTT*H20;",
    "BM10 = KINB*H10/den; BM20 = KTT*H20*BM10/den;",
    "BM30 = KTT*H20*BM20/den; BM40 = KTT*H20*BM30/den;",
    "BM50 = KTT*H20*BM40/den; BM60 = KTT*H20*BM50/den;",
    "BM70 = KTT*H20*BM60/den; BM80 = KTT*H20*BM70/den; BM90 = KTT*H20*BM80/den;",
    "NT0 = BM10+BM20+BM30+BM40+BM50+BM60+BM70+BM80+BM90+NB0;",
    "AC0 = CP0*VD; ADR0 = KSI*NT0*AC0/(KD+CP0);"
  )
  paste0(blk, " ", state, " })")
}

# Order must match modelEquations (ABS, CENT, FRR, DR, B1–B9, NB) for deSolve.
initialConditions = list(
  ABS = 0,
  CENT = gcsf_baseline_ic("AC0+ADR0"),
  FRR = 0,
  DR = 0,
  B1 = gcsf_baseline_ic("BM10"), B2 = gcsf_baseline_ic("BM20"), B3 = gcsf_baseline_ic("BM30"),
  B4 = gcsf_baseline_ic("BM40"), B5 = gcsf_baseline_ic("BM50"), B6 = gcsf_baseline_ic("BM60"),
  B7 = gcsf_baseline_ic("BM70"), B8 = gcsf_baseline_ic("BM80"), B9 = gcsf_baseline_ic("BM90"),
  NB = "NB0"
)
```

# Design

```{r design, echo = TRUE, eval = FALSE}
pk_dense = c(0.167, 0.25, 0.333, 0.5, 0.667, 0.75, 0.833, 1, 1.5, 2, 3, 4, 5,
              6, 8, 10, 12, 14, 16, 18, 20, 24)
anc_dense = c(0.333, 0.5, 0.667, 0.75, 1, 1.5, 2, 3, 4, 5, 6, 8, 10, 12, 14,
               16, 18, 20, 24)
pk_times  = c(pk_dense, pk_dense + 6 * 24, 7 * 24 + c(4, 24, 48))
anc_times = c(anc_dense, (2:6) * 24, anc_dense + 6 * 24, 7 * 24 + c(4, 24, 48, 72))

gcsf_arm = function(dose_mgkg, group = 1L) {
  Arm(
    name = paste0("Group ", group),
    size = 10,
    administrations = list(
      Administration(
        outcome = "ABS",
        timeDose = seq(0, 6 * 24, 24),
        dose = rep(dose_mgkg * 75 * as.numeric(p["FF"]), 7),
        tau = 0
      )
    ),
    samplingTimes = list(
      SamplingTimes(outcome = "RespPK", samplings = pk_times),
      SamplingTimes(outcome = "RespPD", samplings = anc_times)
    ),
    initialConditions = initialConditions
  )
}

design_gcsf = Design(
  name = "gcsf_design",
  arms = list(gcsf_arm(1, 1L), gcsf_arm(3, 2L), gcsf_arm(10, 3L))
)

evaluation_gcsf = Evaluation(
  name = "G-CSF",
  modelEquations = modelEquations,
  modelParameters = modelParameters,
  modelError = modelError,
  outputs = outputs,
  designs = list(design_gcsf),
  fimType = "population",
  odeSolverParameters = list(atol = 1e-8, rtol = 1e-8)
)
```

```{r defs, include = FALSE}
eval(parse(text = knitr::knit_code$get("model-eq")))
eval(parse(text = knitr::knit_code$get("params")))
eval(parse(text = knitr::knit_code$get("design")))
```

# Test plot

PK and ANC profiles for groups 1, 3 and 10 mg/kg (same layout as the
[optimal-design tutorial](https://github.com/timwaterhouse/optimal-design-2023)):

```{r test-plot, echo = TRUE, fig.width = 9, fig.height = 4}
plot_gcsf_test = function(evaluation, n_points = 200) {
  model = defineModelWrapper(defineModelType(evaluation), evaluation)
  arms  = S7::prop(S7::prop(evaluation, "designs")[[1]], "arms")
  times = seq(0, 240, length.out = n_points)

  curves = purrr::list_rbind(purrr::map(seq_along(arms), function(i) {
    plot_arm = arms[[i]]
    S7::prop(plot_arm, "samplingTimes") = list(
      SamplingTimes("RespPK", times),
      SamplingTimes("RespPD", times)
    )
    m = defineModelAdministration(model, plot_arm)
    y = evaluateModel(m, plot_arm)
    rbind(
      data.frame(time = y$RespPK$time, value = y$RespPK$RespPK, model = "PK",  group = i),
      data.frame(time = y$RespPD$time, value = y$RespPD$RespPD, model = "ANC", group = i)
    )
  }))

  samp_pts = purrr::list_rbind(purrr::map(seq_along(arms), function(i) {
    samp = S7::prop(arms[[i]], "samplingTimes")
    rbind(
      data.frame(time = S7::prop(samp[[1]], "samplings"), model = "PK",  group = i),
      data.frame(time = S7::prop(samp[[2]], "samplings"), model = "ANC", group = i)
    )
  }))

  pts = purrr::list_rbind(purrr::map(seq_len(nrow(samp_pts)), function(i) {
    row = samp_pts[i, ]
    sub = curves[curves$model == row$model & curves$group == row$group, ]
    j   = which.min(abs(sub$time - row$time))
    data.frame(time = row$time, value = sub$value[j], model = row$model, group = row$group)
  }))

  ggplot(curves, aes(time, value, color = factor(group))) +
    geom_line(linewidth = 0.9) +
    geom_point(data = pts, aes(time, value), size = 2.5) +
    facet_wrap(~model, scales = "free_y") +
    scale_color_discrete(name = "Group", labels = c("1", "2", "3")) +
    labs(
      x = "Time from first dose (hours)",
      y = "Model predictions",
      title = "Test plot"
    )
}

plot_gcsf_test(evaluation_gcsf)
```

# Evaluate the population FIM

```{r fim, echo = TRUE, eval = FALSE}
# Re-build evaluation_gcsf if needed (see Overview); then:
evaluation_gcsf = run(evaluation_gcsf)
stopifnot(getDeterminant(evaluation_gcsf) > 0)
stopifnot(getRSE(evaluation_gcsf)["μ_NB0", "RSE"] < 50)
show(evaluation_gcsf)
getRSE(evaluation_gcsf)
getSE(evaluation_gcsf)
```

# Compare RSE with PopED (design 1 / 3 / 10 mg/kg)

Reference values from `get_rse(FIM_gcsf, poped_db_gcsf)` in the
[optimal-design-2023](https://github.com/timwaterhouse/optimal-design-2023) tutorial
(same model, sampling grid and `groupsize = 10`).

Run the FIM chunk above first, or in one step:

```r
source("tools/compare-gcsf-rse.R")
```

```{r compare-poped, echo = TRUE, eval = FALSE}
poped_rse = c(
  KA = 3.289140, KEL = 13.518742, VD = 12.604107, KD = 8.265412, KINT = 3.822461,
  KSI = 19.528269, KMT = 4.066511, KTT = 17.017908, NB0 = 10.771361, SC1 = 20.851250,
  SM1 = 8.289988, SM2 = 7.664966,
  w2_KEL = 26.961113, w2_VD = 32.644613, w2_KSI = 31.369081, w2_NB0 = 41.941250,
  w2_SC1 = 29.156946, w2_SM1 = 26.406506,
  sig_slope_PK = 3.900224, sig_inter_PD = 6.541108, sig_slope_PD = 7.713365
)

pfim_rse = getRSE(evaluation_gcsf)
map = c(
  KA = "μ_KA", KEL = "μ_KEL", VD = "μ_VD", KD = "μ_KD", KINT = "μ_KINT",
  KSI = "μ_KSI", KMT = "μ_KMT", KTT = "μ_KTT", NB0 = "μ_NB0", SC1 = "μ_SC1",
  SM1 = "μ_SM1", SM2 = "μ_SM2", w2_KEL = "ω²_KEL", w2_VD = "ω²_VD",
  w2_KSI = "ω²_KSI", w2_NB0 = "ω²_NB0", w2_SC1 = "ω²_SC1", w2_SM1 = "ω²_SM1",
  sig_slope_PK = "σ_slope_RespPK", sig_inter_PD = "σ_inter_RespPD",
  sig_slope_PD = "σ_slope_RespPD"
)

cmp = data.frame(
  parameter = names(poped_rse),
  PopED = unname(poped_rse),
  PFIM = pfim_rse$RSE[unname(map)],
  rel_diff_pct = 100 * abs(pfim_rse$RSE[unname(map)] - poped_rse) / poped_rse,
  row.names = NULL
)

knitr::kable(round(cmp, 2), caption = "RSE (%): PopED vs PFIM (design 1 / 3 / 10 mg/kg)")
```

Typical PFIM vs PopED RSE (same design, after `run(evaluation_gcsf)`):

| Parameter | PopED | PFIM | \|diff\| (%) |
|-----------|------:|-----:|-------------:|
| KA | 3.29 | 1.73 | 47 |
| KEL | 13.52 | 11.84 | 12 |
| VD | 12.60 | 11.75 | 7 |
| KD | 8.27 | 4.41 | 47 |
| KINT | 3.82 | 2.12 | 45 |
| KSI | 19.53 | 14.62 | 25 |
| KMT | 4.07 | 3.59 | 12 |
| KTT | 17.02 | 13.92 | 18 |
| NB0 | 10.77 | 11.21 | 4 |
| SC1 | 20.85 | 19.17 | 8 |
| SM1 | 8.29 | 9.18 | 11 |
| SM2 | 7.66 | 8.56 | 12 |
| ω²_KEL | 26.96 | 29.18 | 8 |
| ω²_VD | 32.64 | 29.79 | 9 |
| ω²_KSI | 31.37 | 34.84 | 11 |
| ω²_NB0 | 41.94 | 27.10 | 35 |
| ω²_SC1 | 29.16 | 28.36 | 3 |
| ω²_SM1 | 26.41 | 26.23 | 1 |
| σ_slope PK | 3.90 | 1.96 | 50 |
| σ_inter PD | 6.54 | 3.05 | 53 |
| σ_slope PD | 7.71 | 13.22 | 71 |

**Agreement:** fixed effects `NB0`, `KEL`, `VD`, `SC1`, `SM1`, `SM2` and most IIV terms (`ω²_KEL`, `ω²_VD`, `ω²_SC1`, `ω²_SM1`) are within about **15%** of PopED.

**Larger gaps:** `KA`, `KD`, `KINT`, `ω²_NB0`, and residual-error RSE reflect differences between PopED's mrgsolve-based FIM and PFIM's finite-difference FIM on this large ODE system — not a singular FIM (`getDeterminant(evaluation_gcsf) > 0`) or a mismatched design.

# PFIM plots after evaluation

```{r pfim-plots, echo = TRUE, eval = FALSE}
plotOptions = list(unitTime = "hour", unitOutcomes = c("ng/mL", "cells/uL"))
plotEvaluation(evaluation_gcsf, plotOptions)
plotRSE(evaluation_gcsf)
plotSE(evaluation_gcsf)
```

# Compare candidate dose sets

```{r compare-doses, echo = TRUE, eval = FALSE}
candidate_doses = list(
  `Design 1` = c(1, 3, 10),
  `Design 2` = c(0.3, 1, 3),
  `Design 3` = c(0.3, 1, 10),
  `Design 4` = c(0.1, 1, 3)
)

purrr::imap(candidate_doses, function(doses, label) {
  d = Design(name = label, arms = purrr::imap(doses, ~ gcsf_arm(.x, .y)))
  ev = Evaluation(
    name = label,
    modelEquations = modelEquations,
    modelParameters = modelParameters,
    modelError = modelError,
    outputs = outputs,
    designs = list(d),
    fimType = "population",
    odeSolverParameters = list(atol = 1e-6, rtol = 1e-6)
  )
  c(design = label, t(round(getRSE(run(ev)), 2)))
})
```
